# Introduction

## Theoretical underpinnings

### Krieger's [-@RN3937, Figure 25] typological concept and the Gary dart point

>...the purpose of a type in archaeology must be to provide an organizational tool which will enable the investigator to group specimens into bodies which have _demonstrable historical meaning in terms of behavior patterns_. Any group which may be labeled a 'type' must embrace material which can be shown to consist of individual variations in the execution of a definite constructional idea; likewise, the dividing lines between a series of types must be based upon demonstrable historical factors, not, as is often the case, upon the inclinations of the analyst or the niceties of descriptive orderliness. [@RN3937, 272]

```{r krieger, fig.fullwidth = TRUE, echo=TRUE, fig.cap = "Step 1: sorting into working patterns of distinct structural plan. Step 2: sorting into detailed working groups each consistently combining features in but one way. Step 3: recombining working groups into tentative types on basis of comparative distributions and associations of working groups. Step 4: consolidation of types through further testing for consistency in form and variation. Step 5: description and illustration of types with full range of variation. Step 6: a - determination of series of linked types, b - determination of material-culture complexes, and c - determination of type relationships in related cultures. This figure is adapted from Krieger (1944:Figure 25)."}
# reconstruct Krieger's graph
library(DiagrammeR)

grViz("
  digraph{
    node[shape = rectangle
         fontname = lato
         penwidth = 0.2]
    Material; Pattern1; Pattern2; Pattern3; 1; 2; 3; 4
    
    node[shape = doublecircle]
    A; B; C; D; E; F; G
    
    node[shape = diamond
         style = filled
         fillcolor = grey]
    x
    
    edge[fontname = lato
         penwidth = 0.2]
    Material->Pattern1; Material->Pattern2; Material->Pattern3;
    Pattern2->A; Pattern2->B; Pattern2->C; Pattern2->D; 
    Pattern2->E; Pattern2->F; Pattern2->G; A->1; C->1; F->1; B->2;
    D->3; E->3; G->4; x->4
    
    graph[nodesep = 0.8]
  }
")
```

### Applying Krieger's model to Gary dart point type-varieties

```{r archepistmod, fig.fullwidth = TRUE, echo=TRUE, fig.cap = "Caption when final."}
# try, fail, try again
grViz("
  digraph{
  
    graph[layout = dot
          rankdir = LR]

    node[shape = doubleoctagon
          fontname = lato
          penwidth = 0.2
          style = filled
          fillcolor = White]
    
    GaryDP
  
    node[shape = doublecircle
         fontname = lato
         penwidth = 0.2
         style = filled
         fillcolor = WhiteSmoke]
    Jtv; PPtv; YMtv; CMtv; GCDtv; Stv
    
    node[shape = egg
         fontname = lato
         style = filled
         fillcolor = White]
    Gs; TyGs; BGs; TGs; SGs; 
    LGs; GLg; GTy; GS; GL; 
    Kf; Al; Ho; Ru; Em; Ke; PM; KN; Cl; 
    G; Ml; LF; Bd; Mn; Cm; 
    G1; G2; 
    Ca; Cb; Cc
    
    node[shape = folder
         style = filled
         fillcolor = Gainsboro]
    Typ; Mal; Lrg; Med; Sml
    
    edge[fontname = lato
         penwidth = 0.2]
    GaryDP->Jtv; GaryDP->PPtv; GaryDP->YMtv; GaryDP->CMtv; GaryDP->GCDtv; GaryDP->Stv;
    Jtv->Gs; Jtv->TyGs; Jtv->BGs; Jtv->TGs; Jtv->SGs; Jtv->LGs;
    PPtv->GLg; PPtv->GTy; PPtv->GS; PPtv->GL; 
    YMtv->Kf; YMtv->Al; YMtv->Ho; YMtv->Ru; YMtv->Em; YMtv->Ke; YMtv->PM; YMtv->KN; YMtv->Cl
    CMtv->G; CMtv->Ml; CMtv->LF; CMtv->Bd; CMtv->Mn; CMtv->Cm;
    GCDtv->G1; GCDtv->G2;
    Stv->Ca; Stv->Cb; Stv->Cc;
    GTy->Typ; Gs->Typ;
    TyGs->Mal; Ml->Mal;
    GLg->Lrg; Ca->Lrg;
    GTy->Med; Cb->Med;
    GS->Sml; Cc->Sml
    
    graph[nodesep = 0.01]
  }
")
```
