# Gary dart point metrics by site

```{r load packages 3, echo=TRUE}
# install required analysis packages
#devtools::install_github("tidyverse/tidyverse")
#devtools::install_github("mlcollyer/RRPP")
#devtools::install_github("kassambara/ggpubr")
#devtools::install_github('sinhrks/ggfortify')
# load libraries
library(tidyverse)
library(RRPP)
library(ggpubr)
library(ggfortify)
library(cluster)
library(wesanderson)
```

## Set working directory, load data, and define variables

```{r setwd ld define variables 3, echo=TRUE}
# set working directory
setwd(getwd())
data<-read.csv("garymorphlm.csv",header = TRUE, as.is=TRUE)
# define variables
maxl<-data$maxl # maximum length
maxw<-data$maxw # maximum width
maxth<-data$maxth # maximum thickness
maxstl<-data$maxstl # maximum stem length
maxstw<-data$maxstw # maximum stem width
site<-data$Site # site name
```

### Calculate dart-arrow index [@RN20881]

Others have noted the gradual diminution in the size of Gary dart points through time [@RN20896;@RN20898;@RN3132;@RN20874], thus it was necessary to begin by asking whether the size of those points aggregated for this study may fall below the arbitrary 11.8 mm dart-arrow index (DAI) threshold posited by Hildebrandt and King [-@RN20881]. The temporal span associated with morphologically-diagnostic Gary dart points is thought to transcend the cultural shift in subsistence practices from hunter-gatherer to emergent horticulturalist, and overlaps temporally with the advent and flourescence of the bow and arrow.

Stem width (neck width) and thickness are among those variables included in this study, and can be used to identify whether Gary dart points may transcend the threshold between dart and arrow points [@RN20881]. The dart-arrow index was calculated for all specimens of Gary dart points used in this study, and none fall below the dart-arrow threshold posited by Hildebrandt and King [-@RN20881]. It is noteworthy that others [@RN20880] have found that the DAI may not be universally applicable; however, the DAI and the threshold identified by Hildebrandt and King [-@RN20881] is a useful heuristic in this instance.

```{r dart-arrow, echo=TRUE}
# calculate dart-arrow index (max stem width + max thickness)
dai<-maxstw + maxth
dai
# how many observations fall at or below 11.8 mm
arrow<-sum(data$dai <= 11.8)
arrow
```

## Boxplots for _variable_ by _site_

```{r boxsite, out.width = "100%", dpi = 300, echo=TRUE, fig.cap = "Boxplots for maximum length, width, thickness, stem length, and stem width for Gary dart points from the Cooper, Means, and Poverty Point sites used in this study."}
# boxplot of maximum length ~ site
sitemaxl<-ggplot(data,aes(x=Site,y=maxl,color=Site)) + geom_boxplot(notch = TRUE) +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3) +
  scale_colour_manual(values = wes_palette("Moonrise2")) +
  theme(legend.position = "none") +
  labs(x = 'Site', y = 'Max Length')
# boxplot of maximum width ~ site
sitemaxw<-ggplot(data,aes(x=Site,y=maxw,color=Site)) + geom_boxplot(notch = TRUE) +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3) +
  scale_colour_manual(values = wes_palette("Moonrise2")) +
  theme(legend.position = "none") +
  labs(x = 'Site', y = 'Max Width')
# boxplot of maximum thickness ~ site
sitemaxth<-ggplot(data,aes(x=Site,y=maxth,color=Site)) + geom_boxplot(notch = TRUE) +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3) +
  scale_colour_manual(values = wes_palette("Moonrise2")) +
  theme(legend.position = "none") +
  labs(x = 'Site', y = 'Max Thick')
# boxplot of stem length ~ site
sitemaxstl<-ggplot(data,aes(x=Site,y=maxstl,color=Site)) + geom_boxplot(notch = TRUE) +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3) +
  scale_colour_manual(values = wes_palette("Moonrise2")) +
  theme(legend.position = "none") +
  labs(x = 'Site', y = 'Stem Length')
# boxplot of stem width ~ site
sitemaxstw<-ggplot(data,aes(x=Site,y=maxstw,color=Site)) + geom_boxplot(notch = TRUE) +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3) +
  scale_colour_manual(values = wes_palette("Moonrise2")) +
  theme(legend.position = "none") +
  labs(x = 'Site', y = 'Stem Width')
# render figure
sitefigure<-ggarrange(sitemaxl,sitemaxw,sitemaxth,sitemaxstl,sitemaxstw,
                  labels = c("a","b","c","d","e"),
                  ncol = 3, nrow = 2)
# plot figure
sitefigure
```

## Principal Components Analysis for _site_

```{r pcasite, out.width = "100%", dpi = 300, echo=TRUE, fig.cap = "PCA by site."}
#attributes for plot
df<-data[c(2:6)]
pch.gps.gp <- c(15,17,19)[as.factor(site)]
col.gps.gp <- wes_palette("Moonrise2")[as.factor(site)]

#pca
autoplot(prcomp(df),
         data = data,
         shape = pch.gps.gp,
         colour = 'Site',
         loadings = TRUE, 
         loadings.colour = 'blue',
         loadings.label = TRUE,
         loadings.label.size = 3, 
         frame = TRUE)+
    scale_fill_manual(values = wes_palette("Moonrise2")) +
    scale_colour_manual(values = wes_palette("Moonrise2"))
```

## Density plots for _variable_ ~ _group_

```{r anova.dens.maxl, out.width = "100%", dpi = 300, echo=TRUE}
# density plot for maxl by site
ggplot(data, aes(x = maxl, fill = Site, colour = Site)) +
  geom_density(alpha = .5) +
  scale_fill_manual(values = wes_palette("Moonrise2")) +
  scale_colour_manual(values = wes_palette("Moonrise2")) +
  xlim(0,150) +
  labs(x = 'Max Length', y = 'density')
```

```{r anova.dens.maxw, out.width = "100%", dpi = 300, echo=TRUE}
# density plot for maxw by site
ggplot(data, aes(x=maxw, fill = Site, colour = Site)) +
  geom_density(alpha = .4) +
  scale_fill_manual(values = wes_palette("Moonrise2")) +
  scale_colour_manual(values = wes_palette("Moonrise2")) +
  xlim(0,75) +
  labs(x = 'Max Width', y = 'density')
```

```{r anova.dens.maxth, out.width = "100%", dpi = 300, echo=TRUE}
# density plot for maxth by site
ggplot(data, aes(x=maxth, fill = Site, colour = Site)) +
  geom_density(alpha = .4) +
  scale_fill_manual(values = wes_palette("Moonrise2")) +
  scale_colour_manual(values = wes_palette("Moonrise2")) +
  xlim(0,25) +
  labs(x = 'Max Thickness', y = 'density')
```

```{r anova.dens.maxstl, out.width = "100%", dpi = 300, echo=TRUE}
# density plot for maxstl by site
ggplot(data, aes(x=maxstl, fill = Site, colour = Site)) +
  geom_density(alpha = .4) +
  scale_fill_manual(values = wes_palette("Moonrise2")) +
  scale_colour_manual(values = wes_palette("Moonrise2")) +
  xlim(0,35) +
  labs(x = 'Max Stem Length', y = 'density')
```

```{r anova.dens.maxstw, out.width = "100%", dpi = 300, echo=TRUE}
# density plot for maxstw by site
ggplot(data, aes(x=maxstw, fill = Site, colour = Site)) +
  geom_density(alpha = .4) +
  scale_fill_manual(values = wes_palette("Moonrise2")) +
  scale_colour_manual(values = wes_palette("Moonrise2")) +
  xlim(0,40) +
  labs(x = 'Max Stem Width', y = 'density')
```
## Analyses of Variance (ANOVA) for _variable_ ~ _site_

### _Maximum length_ of Gary points

```{r anovasitemaxl, out.width = "100%", dpi = 300, echo=TRUE}
# anova = maximum length ~ site
siteml<-lm.rrpp(maxl ~ Site, SS.type = "I",data = data,iter = 9999,print.progress = FALSE)
anova(siteml)

# visualise model predictions for maximum length ~ site
siteDF<-data.frame(Site = c("Cooper","Means","Pov Pt"))
rownames(siteDF)<-c("Cooper","Means","Pov Pt")
sitePreds<-predict(siteml,siteDF)
sitePreds
plot(sitePreds, col = 1:NROW(siteDF))

# pairwise comparison of LS means = which sites differ?
site.pwml<-pairwise(siteml, groups = data$Site)
summary(site.pwml, confidence = 0.95, test.type = "dist")

# pairwise distances between variances = standardization?
summary(site.pwml, confidence = 0.95, test.type = "var")
```

### _Maximum width_ of Gary points

```{r anovasitemaxw, out.width = "100%", dpi = 300, echo=TRUE}
# anova = maximum width ~ site
sitemw<-lm.rrpp(maxw ~ Site, SS.type = "I",data = data,iter = 9999,print.progress = FALSE)
anova(sitemw)

# visualise model predictions for maximum width ~ site
sitePreds<-predict(sitemw,siteDF)
sitePreds
plot(sitePreds, col = 1:NROW(siteDF))

# pairwise comparison = which sites differ?
site.pwmw<-pairwise(siteml, groups = data$Site)
summary(site.pwmw, confidence = 0.95, test.type = "dist")

# pairwise distances between variances = standardization?
summary(site.pwmw, confidence = 0.95, test.type = "var")
```

### _Maximum thickness_ of Gary points

```{r anovasitemaxth, out.width = "100%", dpi = 300, echo=TRUE}
# anova = maximum thickness ~ site
sitemth<-lm.rrpp(maxth ~ Site, SS.type = "I",data = data,iter = 9999,print.progress = FALSE)
anova(sitemth)

# visualise model predictions for maximum thickness ~ site
sitePreds<-predict(sitemth,siteDF)
sitePreds
plot(sitePreds, col = 1:NROW(siteDF))

# pairwise comparison = which sites differ?
site.pwmth<-pairwise(siteml, groups = data$Site)
summary(site.pwmth, confidence = 0.95, test.type = "dist")

# pairwise distances between variances = standardization?
summary(site.pwmth, confidence = 0.95, test.type = "var")
```

### _Maximum stem length_ of Gary points

```{r anovasitemaxstl, out.width = "100%", dpi = 300, echo=TRUE}
# anova = maximum stem length ~ site
sitemstl<-lm.rrpp(maxstl ~ Site, SS.type = "I",data = data,iter = 9999,print.progress = FALSE)
anova(sitemstl)

# visualise model predictions for maximum stem length ~ site
sitePreds<-predict(sitemstl,siteDF)
sitePreds
plot(sitePreds, col = 1:NROW(siteDF))

# pairwise comparison = which sites differ?
site.pwmstl<-pairwise(siteml, groups = data$Site)
summary(site.pwmstl, confidence = 0.95, test.type = "dist")

# pairwise distances between variances = standardization?
summary(site.pwmstl, confidence = 0.95, test.type = "var")
```

### _Maximum stem width_ of Gary points

```{r anovasitemaxstw, out.width = "100%", dpi = 300, echo=TRUE}
# anova = maximum stem width ~ site
sitemstw<-lm.rrpp(maxstw ~ Site, SS.type = "I",data = data,iter = 9999,print.progress = FALSE)
anova(sitemstw)

# visualise model predictions for maximum stem width ~ site
sitePreds<-predict(sitemstw,siteDF)
sitePreds
plot(sitePreds, col = 1:NROW(siteDF))

# pairwise comparison = which sites differ?
site.pwmstw<-pairwise(siteml, groups = data$Site)
summary(site.pwmstw, confidence = 0.95, test.type = "dist")

# pairwise distances between variances = standardization?
summary(site.pwmstw, confidence = 0.95, test.type = "var")
```
