# Poverty Point type-varieties

```{r load packages 4, echo=TRUE}
# install required analysis packages
#devtools::install_github("tidyverse/tidyverse")
#devtools::install_github("rich-iannone/DiagrammeR")
#devtools::install_github("vqv/ggbiplot")
#devtools::install_github("mlcollyer/RRPP")
#devtools::install_github("kassambara/ggpubr")
#devtools::install_github('sinhrks/ggfortify')
# load libraries
library(tidyverse)
library(DiagrammeR)
library(ggbiplot)
library(RRPP)
library(ggpubr)
library(ggfortify)
library(cluster)
```

## Gary varieties proposed by Ford and Webb [-@RN20898]

The `tv1` argument used within this chapter articulates with three variants of the Gary type that include _Large_, _Medium/Typical_, and _Small_ that were assigned using those morphological criteria first advanced by Ford, Phillips, and Haag [-@RN20899] at the Jaketown site, and later refined by Ford and Webb [-@RN20898] at Poverty Point.

```{r setwdld4, echo=TRUE}
# set working directory
setwd(getwd())
# load data
data<-read.csv("garymorphlm.csv",header = TRUE, as.is=TRUE)
```

Each of the three type-varieties proposed by Ford and Webb [-@RN20898] was based upon a suite of morphological criteria that can be systematically replicated, and a function was used to assign each of the Gary points to the correct type-variety. Gary type varieties range between 80 and 140 mm in maxl, and between 30 and 42 mm in maxw for _Gary Large_; between 46 to 79 mm in maxl. Ford and Webb [-@RN20898] also noted ranges in width and thickness values, which they did not use to make their assignments. Those range between 20 and 45 mm in maxw for _Gary Medium/Typical_; and between 33 to 45 mm in maxl, 19 to 32 mm in maxw, and five to 10 mm in maxth for _Gary Small_ [@RN20898].

```{r fordwebbgantt, out.width = "100%", dpi = 300, echo=TRUE,fig.cap = "Gantt charts illustrating the range of linear measurements for each type-variety reported by Ford and Webb (1956)."}
# Ford and Webb (1956)
fwebbl<-data.frame(Name=c('var.GaryLarge','var.GaryMed/Typical','var.GarySmall'),
           Length=c(80,46,30), # in mm (Gary Small changed from 33 to 30)
           end=c(140,80,45) # in mm
)
fwebblength<-ggplot(fwebbl,aes(x=Length,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none")
fwebbw<-data.frame(Name=c('var.GaryLarge','var.GaryMed/Typical','var.GarySmall'),
           Width=c(30,20,19), # in mm
           end=c(42,45,32) # in mm
)
fwebbwidth<-ggplot(fwebbw,aes(x=Width,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none")
# thickness values were not precisely reported for GaryLarge and GaryTypical, and are included here for reference only
fwebbth<-data.frame(Name=c('var.GaryLarge','var.GaryMed/Typical','var.GarySmall'),
           Thickness=c(12.9,9,5), # in mm (actual thickness of var. GaryLarge is 13. Plotted as 12.9-13.1 so that it shows on chart)
           end=c(13.1,10,10) # in mm
)
fwebbthickness<-ggplot(fwebbth,aes(x=Thickness,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none")
# render figure
fwebbfig<-ggarrange(fwebblength,fwebbwidth,fwebbthickness,
                  labels = c("a","b","c"),
                  ncol = 1, nrow = 3)
fwebbfig
```

## Assign Gary type-varieties proposed by Ford and Webb [-@RN20898]

Ford and Webb [-@RN20898] used a single metric (maximum length) to assign Gary varieties at the Poverty Point site. The approach empoyed here enlists all of Ford and Webb's [-@RN20898] reported metrics, including maximum length and maximum width for all varieties, coupled with maximum thickness for  _Gary Small_. No metrics were reported for _Gary Long_, which was omitted from this analysis. Those points that do not match with one of the three criteria were assigned as Gary no variety (Gnv), and, for the purpose of this undertaking, are considered to be Gary dart points with no type-variety assignment.

```{r tv1, echo=TRUE}
# subset poverty point data
pvptmx<-subset(data,Site=="Pov Pt",select=maxl:maxstw)

# assign varieties based upon reported metrics in Ford and Webb [1956]
lcheck1 <- pvptmx$maxl >= 80 & pvptmx$maxl <= 140
lcheck2 <- pvptmx$maxl >= 46 & pvptmx$maxl < 80
lcheck3 <- pvptmx$maxl >= 30 & pvptmx$maxl < 46

pvptmx$tv1 <- "Gnv" # = Gary points (no variety)
pvptmx$tv1 <- ifelse(lcheck1, "L", pvptmx$tv1)
pvptmx$tv1 <- ifelse(lcheck2, "M/T", pvptmx$tv1)
pvptmx$tv1 <- ifelse(lcheck3, "S", pvptmx$tv1)

tv1<-pvptmx$tv1
tv1
```

## Boxplots for Ford and Webb's [-@RN20898] type-varieties of Gary dart points from Poverty Point

```{r boxplotpp, out.width = "100%", dpi = 300, echo=TRUE, fig.cap = "Boxplots for a, maximum length; b, width; c, thickness; d, stem length; and e, stem width for Gary dart points from the Poverty Point site."}

# boxplot of maximum length
pvptmaxl<-ggplot(pvptmx,aes(x=tv1,y=maxl,color=tv1)) + 
  geom_boxplot() +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none")
# boxplot of maximum width
pvptmaxw<-ggplot(pvptmx,aes(x=tv1,y=maxw,color=tv1)) + 
  geom_boxplot() +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3)+
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none")
# boxplot of maximum thickness
pvptmaxth<-ggplot(pvptmx,aes(x=tv1,y=maxth,color=tv1)) + 
  geom_boxplot() +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3)+
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none")
# boxplot of maximum stem length
pvptmaxstl<-ggplot(pvptmx,aes(x=tv1,y=maxstl,color=tv1)) + 
  geom_boxplot() +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3)+
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none")
# boxplot of maximum thickness
pvptmaxstw<-ggplot(pvptmx,aes(x=tv1,y=maxstw,color=tv1)) + 
  geom_boxplot() +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3)+
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none")
# render figure
ppfigure<-ggarrange(pvptmaxl,pvptmaxw,pvptmaxth,pvptmaxstl,pvptmaxstw,
                  labels = c("a","b","c","d","e"),
                  ncol = 3, nrow = 2)
ppfigure
```

## Density plots for variables by type-varieties

### Density plot for maximum length by tv1

```{r ppdensity1, out.width = "100%", dpi = 300, echo=TRUE}
# define variables in poverty point subset
pp.maxl<-pvptmx$maxl
pp.maxw<-pvptmx$maxw
pp.maxth<-pvptmx$maxth
pp.maxstl<-pvptmx$maxstl
pp.maxstw<-pvptmx$maxstw
pp.tv1<-pvptmx$tv1
# density plot of maxl by tv1
ggplot(pvptmx, aes(x=pp.maxl, fill = pp.tv1)) +
  geom_density(alpha = .4) +
  scale_color_brewer(palette = "Set1") +
  xlim(30,140) #full range of variation included in the linear measurements associated with type-varieties
```

### Density plot for maximum width by tv1

```{r ppdensityw, out.width = "100%", dpi = 300, echo=TRUE}
# density plot of maxw by tv1
ggplot(pvptmx, aes(x=pp.maxw, fill = pp.tv1)) +
  geom_density(alpha = .4) +
  scale_color_brewer(palette = "Set1") +
  xlim(5,45)
```

### Density plot for maximum thickness by tv1

```{r ppdensityth, out.width = "100%", dpi = 300, echo=TRUE}
# density plot of maxth by tv1
ggplot(pvptmx, aes(x=pp.maxth, fill = pp.tv1)) +
  geom_density(alpha = .4) +
  scale_color_brewer(palette = "Set1") +
  xlim(0,30)
```

### Density plot for maximum width by tv1

```{r ppdensitystl, out.width = "100%", dpi = 300, echo=TRUE}
# density plot of maxstl by tv1
ggplot(pvptmx, aes(x=pp.maxstl, fill = pp.tv1)) +
  geom_density(alpha = .4) +
  scale_color_brewer(palette = "Set1") +
  xlim(0,30)
```

### Density plot for maximum width by tv1

```{r ppdensitystw, out.width = "100%", dpi = 300, echo=TRUE}
# density plot of maxstw by tv1
ggplot(pvptmx, aes(x=pp.maxstw, fill = pp.tv1, colour = pp.tv1)) +
  geom_density(alpha = .4) +
  scale_color_brewer(palette = "Set1") +
  xlim(0,40)
```

## Principal Components Analysis for Ford and Webb's [-@RN20898] type-varieties at Poverty Point

```{r pca, out.width = "100%", dpi = 300, echo=TRUE, fig.cap = "PCA by Gary type-varieties reported by Ford and Webb (1956) at Poverty Point."}
#pca
df<-pvptmx[c(1:5)]
t1pca<-autoplot(prcomp(df),data = pvptmx, colour = 'tv1',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE,loadings.label.size = 3, 
         frame = TRUE)
#render figure
t1pca
```

## Analyses of Variance (ANOVA) for _linear variables_ ~ _type-varieties_

### _Maximum length_ ~ _type-variety_ at Poverty Point

```{r pp.maxl-x-tv, out.width = "100%", dpi = 300, echo=TRUE}
# anova = maximum length ~ tv1
t1ml<-lm.rrpp(pp.maxl ~ pp.tv1, SS.type = "I",data = pvptmx,iter = 9999,print.progress = FALSE)
anova(t1ml)
# visualise model predictions for maximum length ~ type-variety at poverty point
pp.tvDF<-data.frame(pp.tv1 = c("L","M/T","S"))
rownames(pp.tvDF)<-c("L","M/T","S")
ppPreds<-predict(t1ml,pp.tvDF)
ppPreds
plot(ppPreds, col = 1:NROW(pp.tvDF))
# pairwise comparison of LS means = which sites differ?
pp.ml<-pairwise(t1ml, groups = pvptmx$tv1)
summary(pp.ml, confidence = 0.95, test.type = "dist")
# correlation between mean vectors (angles in degrees)
summary(pp.ml, confidence = 0.95, test.type = "VC", angle.type = "deg")
# pairwise distances between variances = standardization?
summary(pp.ml, confidence = 0.95, test.type = "var")
```

### _Maximum width_ ~ _type-variety_ at Poverty Point

```{r pp.maxw-x-tv, out.width = "100%", dpi = 300, echo=TRUE}
# anova = maximum width ~ tv1
t1mw<-lm.rrpp(pp.maxw ~ pp.tv1, SS.type = "I",data = pvptmx,iter = 9999,print.progress = FALSE)
anova(t1mw)
# visualise model predictions for maximum width ~ type-variety at poverty point
ppPreds<-predict(t1mw,pp.tvDF)
ppPreds
plot(ppPreds, col = 1:NROW(pp.tvDF))
```

### _Maximum thickness_ ~ _type-variety_ at Poverty Point

```{r pp.maxth-x-tv, out.width = "100%", dpi = 300, echo=TRUE}
# anova = maximum thickness ~ tv1
t1mth<-lm.rrpp(pp.maxth ~ pp.tv1, SS.type = "I",data = pvptmx,iter = 9999,print.progress = FALSE)
anova(t1mth)
# visualise model predictions for maximum thickness ~ type-variety at poverty point
ppPreds<-predict(t1mth,pp.tvDF)
ppPreds
plot(ppPreds, col = 1:NROW(pp.tvDF))
```

### _Maximum stem length_ ~ _type-variety_ at Poverty Point

```{r pp.maxstl-x-tv, out.width = "100%", dpi = 300, echo=TRUE}
# anova = maximum stem length ~ tv1
t1mstl<-lm.rrpp(pp.maxstl ~ pp.tv1, SS.type = "I",data = pvptmx,iter = 9999,print.progress = FALSE)
anova(t1mstl)
# visualise model predictions for maximum stem length ~ type-variety at poverty point
ppPreds<-predict(t1mstl,pp.tvDF)
ppPreds
plot(ppPreds, col = 1:NROW(pp.tvDF))
```

### _Maximum stem width_ ~ _type-variety_ at Poverty Point

```{r pp.maxstw-x-tv, out.width = "100%", dpi = 300, echo=TRUE}
# anova = maximum stem width ~ tv1
t1mstw<-lm.rrpp(pp.maxstw ~ pp.tv1, SS.type = "I",data = pvptmx,iter = 9999,print.progress = FALSE)
anova(t1mstw)
# visualise model predictions for maximum stem length ~ type-variety at poverty point
ppPreds<-predict(t1mstw,pp.tvDF)
ppPreds
plot(ppPreds, col = 1:NROW(pp.tvDF))
```
